\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-13 paper.qmd
Update center to centroid
Need to fix the prediction fig

A new benefit of this fitted model is that it allows us to now predict a new observation's value in the NLDR, for any method. The steps are to determine the closest bin center in \pD{}, $C^{(p)}_{h}$ and predict it to be the centroid of this bin in \gD{}, $C^{(2)}_{h}$. This can be written as, let $z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}$, then the new observation $i$ falls in the hexagon, $H_h = \{i| z(i) = h\}$ and the corresponding \kD{} bin centroids, $C_{h}^{(2)} = (c_{h1}, c_{h2})$.

@@

>>>>

A new benefit of this fitted model is that it allows us to now predict a new observation's value in the NLDR, for any method. The steps are to determine the closest bin centroid in \pD{}, $C^{(p)}_{h}$ and predict it to be the centroid of this bin in \gD{}, $C^{(2)}_{h}$. This can be written as, let $z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}$, then the new observation $i$ falls in the hexagon, $H_h = \{i| z(i) = h\}$ and the corresponding \kD{} bin centroids, $C_{h}^{(2)} = (c_{h1}, c_{h2})$.

@@

```{r}
#| label: model-prediction-umap-original

predict_umap_scurve <- read_rds(file = "data/s_curve/s_curve_umap_predict_true.rds")

predict_scurve_obj <- gen_scaled_data(
    data = predict_umap_scurve)

predict_umap_scurve_scaled <- predict_scurve_obj$scaled_nldr

plot_predict_umap <- ggplot(data = predict_umap_scurve_scaled,
                            aes(
                              x = UMAP1,
                              y = UMAP2
                            )) +
  geom_point()
```

```{r}
#| label: model-prediction-umap-quollr

true_model_df <- read_rds("data/s_curve/scurve_true_model.rds")

# Initialize effective bins along x
effective_bin1_scurve <- 15

# Fit high-dimensional model
scurve_model <- fit_highd_model(
  training_data = training_data_scurve,
  emb_df = umap_scurve_scaled,
  bin1 = effective_bin1_scurve,
  r2 = r2,
  q = 0.16,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

true_pred_df <- predict_emb(
  test_data = true_model_df,
  df_bin_centroids = df_bin_centroids_scurve,
  df_bin = df_bin_scurve,
  type_NLDR = "UMAP"
)

# Compute radius r
a1 <- calc_bins_y(bin1 = 15, r2 = r2, q = 0.16)$a1
r <- a1 / 2

# Function to jitter points within a circumcircle of a hexagon
jitter_within_circumcircle <- function(center_x, center_y, radius, num_points) {
  theta <- runif(num_points, 0, 2 * pi)
  r <- radius * sqrt(runif(num_points))
  x <- center_x + r * cos(theta)
  y <- center_y + r * sin(theta)
  jittered_points <- data.frame(UMAP1 = x, UMAP2 = y)
  return(jittered_points)
}

# Jittered points data frame
jittered_points_df <- data.frame()

# Iterate through each hexagon and jitter points within the circumcircle
for (hex_id in unique(hex_grid_nonempty$hex_poly_id)) {
  hex_points <- hex_grid_nonempty %>% filter(hex_poly_id == hex_id)
  center_x <- mean(hex_points$x)
  center_y <- mean(hex_points$y)
  points_in_hex <- true_pred_df %>% filter(pred_UMAP_1 >= min(hex_points$x) & pred_UMAP_1 <= max(hex_points$x) & pred_UMAP_2 >= min(hex_points$y) & pred_UMAP_2 <= max(hex_points$y))

  if (nrow(points_in_hex) > 0) {
    jittered_points <- jitter_within_circumcircle(center_x, center_y, r, nrow(points_in_hex))
    jittered_points_df <- rbind(jittered_points_df, jittered_points)
  }
}

plot_predict_umap_model <- ggplot(data = jittered_points_df,
                            aes(
                              x = UMAP1,
                              y = UMAP2
                            )) +
  geom_point()
```

```{r}
#| label: model-prediction-points-line

predict_umap_scurve_scaled <- predict_umap_scurve_scaled |>
  mutate(type = "predict_from_umap") |>
  mutate(point_id = row_number())

jittered_points_df <- jittered_points_df |>
  mutate(type = "predict_from_model") |>
  mutate(point_id = row_number())

predict_point_df <- bind_rows(
  predict_umap_scurve_scaled,
  jittered_points_df
)

plot_predict_umap_model_connect <- ggplot(data = predict_point_df,
                            aes(
                              x = UMAP1,
                              y = UMAP2, 
                              group = point_id
                            )) +
  geom_point() +
  geom_line()
```

```{r}
#| echo: false
#| label: fig-predict-scurve
#| fig-pos: H
#| fig-width: 9
#| fig-height: 5
plot_predict_umap + plot_predict_umap_model + plot_predict_umap_model_connect 
```

### 2024-08-13 paper.tex
Update center to centroid
Need to fix the prediction fig

A new benefit of this fitted model is that it allows us to now predict a
new observation's value in the NLDR, for any method. The steps are to
determine the closest bin center in \pD{}, \(C^{(p)}_{h}\) and predict
it to be the centroid of this bin in \gD{}, \(C^{(2)}_{h}\). This can be
written as, let
\(z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}\),
then the new observation \(i\) falls in the hexagon,
\(H_h = \{i| z(i) = h\}\) and the corresponding \kD{} bin centroids,
\(C_{h}^{(2)} = (c_{h1}, c_{h2})\).

>>>>

A new benefit of this fitted model is that it allows us to now predict a
new observation's value in the NLDR, for any method. The steps are to
determine the closest bin centroid in \pD{}, \(C^{(p)}_{h}\) and predict
it to be the centroid of this bin in \gD{}, \(C^{(2)}_{h}\). This can be
written as, let
\(z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}\),
then the new observation \(i\) falls in the hexagon,
\(H_h = \{i| z(i) = h\}\) and the corresponding \kD{} bin centroids,
\(C_{h}^{(2)} = (c_{h1}, c_{h2})\).

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-predict-scurve-1.pdf}

\caption{\label{fig-predict-scurve}}
